---
title: "p8105_ggplot_pt1"
author: "Johnstone Tcheou"
date: "2024-09-26"
output: html_document
---

# `ggplot` pt 1

## Why `ggplot`

Simplifies graphics to a system that can be applied to other graphics systems - if you know `ggplot`, you can get your head around `leaflet` and other visualization packages

## Learning `ggplot`

There are some things you will need to know and some things you can search up

## Using ggplot

Focusing on first part of basic graph components today

- Data (always start with dataframe first)
- Aesthetic mappings (variables in dataframe will map to specific aesthetics - axes, density, color, etc)
- `geoms` (what kind of a graph are you actually trying to make - what are the geometries you want to include)

Advanced graph components include

- Facets
- Scales
- Statistics

Graphs are built by combining these components
Components are consistent across graph types - scatterplots, bar graphs, density plots, ridge pltos 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggridges)
```

## `rnoaa` import data

`rnoaa` is a way to interact with the data that they make public 
This code chunk pulls the data, then `mutate`  to create a variable more intuitive than `id` - their actual locations, divide `tmin` and `tmax` by 10, and then just reorders the variables with `select` so `name` and `id` come first 

```{r rnoaa import}
weather_df <- 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = case_match(
      id, 
      "USW00094728" ~ "CentralPark_NY", 
      "USW00022534" ~ "Molokai_HI",
      "USS0023B17S" ~ "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())
```

## Making our first plot

Define dataset
Define aesthetic mappings
Then define geometries afterwards

Might want to run the `ggplot` function with just the x and y axes specified to ensure that our scales are as what we intend it to 

Both function calls below work, but sometimes you may prefer the latter structure in case you want to do something to data like `filter` and then pipe it into `ggplot` once you've done some data manipulation, as is common

You can also save `ggplot` objects to a variable - call it to display it 

If you want your plot images to show up on GitHub, named after their chunks accordingly, will need to commit

Note that whenever running `ggplot`, error message that **17 rows removed containing missing values or values outside scale range** 

Can check for rows with missing values for `tmin` or `tmax` 

```{r first_plot}
ggplot(weather_df, aes(x=tmin, y=tmax, color=name)) +
  geom_point()

weather_df |>
  ggplot(aes(x=tmin, y=tmax, color=name)) +
  geom_point()

weather_df |>
  filter(is.na(tmin) | is.na(tmax))
```

## More complicated graphs

`alpha` option designates shading/transparency
`size` makes things smaller
`geom_smooth` creates a smooth curve regression line through data, where `se` option is the standard error calculated in the curve - `FALSE` turns it off
**Where you place your aesthetics matter** - declaring the aesthetic in one geometry alone is restricted to that geometry 
e.g. `geom_smooth` no longer gets color by name, it is just one color
If you do not get the output you intended, probably an **aesthetic mismatch** somewhere

```{r complicated_graphs}
weather_df |>
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(se=FALSE)


weather_df |>
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point(aes(color = name), alpha = 0.3, size = 0.8) +
  geom_smooth(se=FALSE)
```

## Faceting 

**Faceting** lets us 
If points are overlapping too much, we can separate into scatterplot for each one with `facet_grid`
`~` in `facet_grid` specifies what rows are on left side and what columns are in the grid on right side 
`.` designates nothing in that side
`.` in rows puts scatterplots in columns first, `.` in columns puts scatterplots in rows first
Columns first is preferred for Jeff, but comes down to personal preference

```{r faceting}
weather_df |>
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(se=FALSE) + 
  facet_grid(. ~ name)

weather_df |>
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(se=FALSE) + 
  facet_grid(name ~ .)

```

## More interesting scatterplot

Add aesthetic so size is based on precipitation level with `size` option, where more precipitation = larger circle, less precipitation = smaller circle 
Warm in Seattle during summer and no rain, but in winter, cooler and tons of rain

```{r interesting_scatterplot}
weather_df |>
  ggplot(aes(x = date, y = tmax, color = name, size = prcp)) + 
  geom_point(alpha = 0.3) + 
  geom_smooth(se = FALSE) + 
  facet_grid(. ~ name)

weather_df |>
  filter(name == "CentralPark_NY") |>
  ggplot(aes(x = date, y = tmax, size = prcp)) + 
  geom_point(alpha = 0.3) + 
  geom_smooth(se = FALSE) 

```

## Learning assessment

Shows that we don't want to overwrite and change up original data, sometimes we just want to visualize small portions or components of the data 
Hence, we can have a pipe like this that doesn't overwrite the original dataset

```{r learning_assessment}
weather_df |>
  filter(name == "CentralPark_NY") |>
  mutate(
    tmax_f = tmax * 9/5 + 32,
    tmin_f = tmin * 9/5 + 32
  ) |>
  ggplot(aes(x = tmin_f, y = tmax_f)) + 
  geom_point(alpha = 0.3) + 
  geom_smooth(se = FALSE) 
```

## geom_smooth

`?geom_smooth` to get details on `method`, where we can change the smoothing method from `loess` default to `lm` for linear model, giving us straight line instead

```{r geom_smooth_method}
weather_df |>
  filter(name == "CentralPark_NY") |>
  mutate(
    tmax_f = tmax * 9/5 + 32,
    tmin_f = tmin * 9/5 + 32
  ) |>
  ggplot(aes(x = tmin_f, y = tmax_f)) + 
  geom_point(alpha = 0.3) + 
  geom_smooth(se = FALSE) 

weather_df |>
  filter(name == "CentralPark_NY") |>
  mutate(
    tmax_f = tmax * 9/5 + 32,
    tmin_f = tmin * 9/5 + 32
  ) |>
  ggplot(aes(x = tmin_f, y = tmax_f)) + 
  geom_point(alpha = 0.3) + 
  geom_smooth(se = FALSE, method = lm) 
```
## Small things

Can mix and match elements accordingly, like removing the data points but keeping the regression models

Say maybe you run a scatterplot like this but you have tens and tens of thousands of data points - becomes hard to distinguish points from each other
`geom_hex` shows where you have the highest peaks in your data

Can also set all points to one color
`"blue"` sets everything to one color, but `blue` sets everything to red  
What `ggplot` is doing is creating a variable called `"blue"` of all blue things 
When using `blue`, it tries to look for a variable called `blue` which is not set yet, and fails, so it uses red by default

How to know when something needs to be in quotes or not?
If something is a variable name, does not need to be in quotes
If not a variable name, usually needs to be in quotes

```{r all_the}
weather_df |>
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_smooth(se=FALSE)

weather_df |>
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point()

weather_df |>
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_hex()

weather_df |>
  ggplot(aes(x = tmin, y = tmax, color = blue)) +
  geom_point()

weather_df |>
  ggplot(aes(x = tmin, y = tmax, color = "blue")) +
  geom_point()
```

## Univariate plots

Same stuff as every other plot, take data, give aesthetic mapping, tell it geometries to use 
Gets wonky when trying to compare, like when `color` is name, tons of overlapping bars
`position = dodge` ensures that scatterplot bars are next to each other than on top of each other 
Scatterplot is still hard to read but perhaps is more understandable 
Maybe want to use `facet_grid` to separate out each group based on `name`
If you want to compare everything all at once, maybe use different plot type instead of histogram, like a density plot which are like smooth histograms
Histogram just reveals the spread and distribution of our data by each group 
If you want to talk about features of datasets, **distribution is really helpful**
`geom_density` allows us to make more quick and easy comparisons
(also if you want to fill bar instead of coloring outline as `color` does, use `fill`)

```{r histogram}
weather_df |>
  ggplot(aes(x = tmin)) + 
  geom_histogram()

weather_df |>
  ggplot(aes(x = tmin, fill = name)) + 
  geom_histogram()

weather_df |>
  ggplot(aes(x = tmin, fill = name)) + 
  geom_histogram(position = "dodge")

weather_df |>
  ggplot(aes(x = tmin, fill = name)) + 
  geom_histogram(position = "dodge")

weather_df |>
  ggplot(aes(x = tmin, fill = name)) + 
  geom_density(alpha = 0.3)
```

## Box plots

`name` on x axis, `tmin` on y axis
In this case, `color` not all too useful since names are already on x axis, but maybe color helps reinforce the names or you want to maintain color consistency across figures

```{r box}
weather_df |>
  ggplot(aes(x = name, y = tmin, fill = name)) + 
  geom_boxplot()
```

## Violin plots

Violin plots work by taking density plots and flipping them
You get everything from density or ridge plots, but making them vertical
Jeff prefers violin plots over box plots to see things more accurately about distribution like skew, multimodal, etc
Sometimes boxes cannot reflect multimodal nature of some datasets, so violin plots more revealing 

```{r violin}
weather_df |>
  ggplot(aes(x = name, y = tmin, fill = name)) + 
  geom_violin()
```

## Ridge plots

Ridge plots will try to create separate densities for each variable and spread them across on y axis
`ggridges` package is where `geom_density_ridges` are - ensure you have the packages installed and loaded
`geom_density_ridges` provides each density, **separated out and vertically**, for quick and easy visual comparison
Central Park looks bimodal, and Waterhole looks somewhat bimodal 

```{r ridge}
weather_df |>
  ggplot(aes(x = tmin, y = name)) + 
  geom_density_ridges()
```

## Learning assessment

Since it is hard to see things just outside of the tail, 

Box plots show a little more detail about skewness and distribution - per outliers, most days it does not rain
But if we want to look at days where it is raining, filter for days where precipitation is > 0 or > 10
If you have an outlier that is really stretching the axis scale, you can run another visualization that omits the one outlier 

```{r learning_assessment_2}
weather_df |>
  ggplot(aes(x = prcp, fill = name)) + 
  geom_histogram(position = "dodge")

weather_df |>
  ggplot(aes(x = prcp, fill = name)) + 
  geom_density()

weather_df |>
  ggplot(aes(x = name, y = prcp, fill = name)) + 
  geom_boxplot()

weather_df |>
  ggplot(aes(x = name, y = prcp, fill = name)) + 
  geom_violin()

weather_df |>
  ggplot(aes(x = prcp, y = name, fill = name)) + 
  geom_density_ridges()



weather_df |>
  filter(prcp > 10) |>
  ggplot(aes(x = prcp, fill = name)) + 
  geom_density()


```

## Saving and embedding plots

As seen before, `ggsave` lets us save a plot to file format we prefer and name it as we desire
By default, `ggsave` saves the most recent plot 
If we know there's a plot we want to save that might not be the most recent plot, then we can **save that desired plot as a variable, and then use `ggsave` on it
We can also set `width` and `height` in the `ggsave` function
If you know you'll be creating a lot of plots, may be worth making a separate folder for plots and saving your plots there

```{r save}
weather_df |>
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point()

ggsave("ggp_weather.pdf")

ggp_weather <- weather_df |>
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point()

ggsave("plots/ggp_weather.pdf", ggp_weather, width = 8, height = 6)

```

To embed, use chunk options to determine figure width and figure height
In background, R made figure that it tried to fit to the R Markdown format
Then it shrunk down to fit per our specified width and height, which affects how some things are displayed like text
`fig.asp` sets aspect ratio like 0.6, 60% as high as it is wide
We can set some of these chunk options to be global options 

```{r embed, fig.width=12, fig.asp=0.6}
weather_df |>
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point()

```

